\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Yans updates

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

>>>>

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
in this order)
More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.
